---
title: "Improving computational performance with algorithm engineering"
author: "Kirill Müller"
date: "July 1, 2015"
output: ioslides_presentation
---

## Algorithm engineering

> Design, analysis, implementation, optimization, profiling and experimental evaluation of computer algorithms. (*Wikipedia*)

Here: Substitute existing implementation by an algorithm
with a lower computational complexity.

**Examples**

1. Weighted random sampling without replacement
2. Similarity-based statistical matching
3. Survey calibration
4. Package management


## Random sampling

```{r}
sample.int
```

- `replace`: With or without replacement
- `prob`: Uniform or non-uniform probabilities


## Implementation of random sampling

Common framework:

- Subdivide an interval according to probabilities
- Choose sub-interval by sampling a uniform point in interval
- If sampling without replacement, remove sub-interval



## Slide with Bullets

- Bullet 1
- Bullet 2
- Bullet 3

## Slide with R Code and Output

```{r}
summary(cars)
```

## Slide with Plot

```{r, echo=FALSE}
plot(cars)
```

